---
title: "Survey Data Analysis — Report 2"
author: "Michał Marchwiak 276003 Weronika Mitulska 277475"
date: today
lang: en
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    documentclass: article
    geometry:
      - margin=2.2cm
    fontsize: 11pt
    fig-width: 6.5
    fig-height: 4.2
    fig-align: center
    include-in-header:
      text: |
        
        \usepackage{booktabs}
        \usepackage{float}
        \usepackage{amsmath, amssymb}
execute:
  echo: false
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    comment: "#>"
    collapse: true
---

```{r setup, include=FALSE}
# Libraries
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
  library(kableExtra)
  library(DescTools)    # gamma, Kendall tau-b, MultinomCI (Sison-Glaz)
  library(vcd)          # assoc plot
  library(energy)       # dcor (task *1)
})

set.seed(2025)

# Load data (CSV in UTF-8, separator ';')
dane <- read.csv2("ankieta.csv", stringsAsFactors = FALSE, fileEncoding = "UTF-8")
colnames(dane) <- c("DZIAL","STAZ","CZY_KIER","PYT_1","PYT_2","PYT_3","PLEC","WIEK")

# Age categorization: 4 groups
dane$WIEK_KAT <- cut(dane$WIEK,
                     breaks = c(-Inf, 35, 45, 55, Inf),
                     labels = c("up to 35", "36-45", "46-55", "over 55"),
                     right  = TRUE)

# CZY_ZADOW variable defined in report 1:
# employee satisfied (Tak) when PYT_1 > 0, dissatisfied (Nie) otherwise.
dane$CZY_ZADOW <- factor(ifelse(dane$PYT_1 > 0, "Tak", "Nie"),
                         levels = c("Nie","Tak"))

# Ordered response scales (needed for gamma/tau)
dane$PYT_1_ord <- factor(dane$PYT_1, levels = c(-2,-1,0,1,2), ordered = TRUE)
dane$PYT_2_ord <- factor(dane$PYT_2, levels = c(-2,-1,1,2),   ordered = TRUE)
dane$STAZ_ord  <- factor(dane$STAZ,  levels = c(1,2,3),       ordered = TRUE)

# Labels
etyk_pyt1 <- c("very dissat.", "dissat.", "no opinion", "sat.", "very sat.")
etyk_pyt2 <- c("str. disagree","disagree","agree","str. agree")
```

# Introduction

This report continues report 1 on the employee survey
(n = 200, simple random sampling with replacement, four departments: IT, HR, MK, PD).
Variable conventions from the previous report are preserved, in particular
the binary variable `CZY_ZADOW` (Tak if `PYT_1 > 0`) and the
`WIEK_KAT` variable categorizing age into four groups:
`up to 35`, `36--45`, `46--55`, `over 55`.

All computations were performed in R. The random seed was set to `set.seed(2025)`
(for simulation-based tests, including Freeman--Halton).

---

# Part I

## Task 1 — Confidence interval for a probability vector

Niech $X = (X_1,\dots,X_5) \sim \mathrm{Multinom}(n,p)$, gdzie
$p = (p_1,\dots,p_5)$ is the probability vector odpowiedzi
"very dissat.", "dissat.", "no opinion", "sat.", "very sat.".
The estimator is $\hat p_i = X_i/n$.

**Simultaneous confidence interval at level $1-\alpha$** is obtained by two
classical methods:

- **Goodman (1965)** — based on inverted $\chi^2$ with Bonferroni correction
  (quantile $\chi^2_{1,\,\alpha/k}$, where $k$ = number of categories). For each
  component:
  $$
  p_i \in \frac{A + 2 X_i \pm \sqrt{A\,(A + 4 X_i(n-X_i)/n)}}{2(n+A)},\qquad A = \chi^2_{1,\alpha/k}.
  $$

- **Sison, Glaz (1995)** — equal-width interval
  $\hat p_i \pm d$, where $d$ is chosen numerically so that
  simultaneous coverage is $\ge 1-\alpha$.

```{r zad1}
# Data from the task statement
x  <- c(14, 17, 40, 100, 29)
n  <- sum(x)
p_hat <- x / n
alpha <- 0.05
k  <- length(x)

# --- (a) Goodman (Bonferroni + inverted chi^2_1) ---
goodman_ci <- function(x, alpha = 0.05) {
  n <- sum(x); k <- length(x)
  A <- qchisq(1 - alpha/k, df = 1)
  low <- (A + 2*x - sqrt(A*(A + 4*x*(n - x)/n))) / (2*(n + A))
  up  <- (A + 2*x + sqrt(A*(A + 4*x*(n - x)/n))) / (2*(n + A))
  cbind(lower = low, upper = up)
}
ci_g <- goodman_ci(x, alpha)

# --- (b) Sison-Glaz (DescTools::MultinomCI) ---
ci_sg_full <- DescTools::MultinomCI(x, conf.level = 1 - alpha, method = "sisonglaz")
ci_sg <- ci_sg_full[, c("lwr.ci", "upr.ci")]

tab1 <- data.frame(
  Category   = etyk_pyt1,
  Count  = x,
  `p_hat`     = round(p_hat, 4),
  `Goodman_L` = round(ci_g[,1], 4),
  `Goodman_U` = round(ci_g[,2], 4),
  `SG_L`      = round(ci_sg[,1], 4),
  `SG_U`      = round(ci_sg[,2], 4),
  check.names = FALSE
)
kable(tab1, caption = "Confidence intervals 95\\% dla p — metody Goodmana i Sison-Glaz.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

**Table description 1.** Column `p_hat` gives the MLE; subsequent column pairs contain lower and upper bounds of simultaneous 95% CIs. Sison--Glaz intervals are noticeably narrower than Goodman's (S--G is equal-width at the cost of asymptotic accuracy), but substantive conclusions agree: the "satisfied" category dominates ($p_4 \in [0.4133, 0.5906]$ Goodman, $[0.4305,0.5705]$ S--G), while extremely dissatisfied categories ($p_1, p_2$) have probabilities of a few percent with intervals disjoint from the dominant category.

```{r zad1-plot, fig.cap="Point estimates $\\hat p_i$ (points) and simultaneous 95\\% CIs by Goodman (solid) and Sison-Glaz (dashed)."}
df_pl <- data.frame(
  kat = factor(etyk_pyt1, levels = etyk_pyt1),
  p_hat = p_hat,
  gL = ci_g[,1], gU = ci_g[,2],
  sL = ci_sg[,1], sU = ci_sg[,2]
)
ggplot(df_pl, aes(x = kat, y = p_hat)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = gL, ymax = gU), width = 0.18) +
  geom_errorbar(aes(ymin = sL, ymax = sU), width = 0.32, linetype = 2) +
  labs(x = "Response category", y = expression(hat(p))) +
  theme_minimal(base_size = 11)
```

**Figure description 1.** Estimates rise monotonically from the most negative category to "satisfied" and dip slightly for "very satisfied". Goodman CIs are wider for low-count categories (asymptotic effect of inverted $\chi^2_1$).

\newpage

## Task 2 — p-value function in $\chi^2$ tests for $H_0: p = p_0$

Dla obserwacji $x = (x_1,\dots,x_k)$ wektora $X \sim \mathrm{Multinom}(n,p)$
testujemy $H_0: p = p_0$. Statystyki:

$$
T_{\mathrm{P}}(x) = \sum_{i=1}^{k} \frac{(x_i - n p_{0i})^2}{n p_{0i}},
\qquad
T_{\mathrm{NW}}(x) = 2 \sum_{i=1}^{k} x_i \log\!\Bigl(\frac{x_i}{n p_{0i}}\Bigr).
$$

Przy $H_0$ obie statystyki are asymptotically distributed as $\chi^2_{k-1}$,
zatem p-value $= 1 - F_{\chi^2_{k-1}}(T_\cdot(x))$.

```{r zad2}
# Pearson chi-square and LR test for H0: p = p0
p_value_multinom <- function(x, p0,
                             method = c("pearson", "LR"),
                             exact = FALSE, B = 10000) {
  method <- match.arg(method)
  stopifnot(abs(sum(p0) - 1) < 1e-10, length(x) == length(p0), all(x >= 0))
  n <- sum(x); k <- length(x); np0 <- n * p0

  stat <- if (method == "pearson") {
    sum((x - np0)^2 / np0)
  } else {
    # konwencja 0 log 0 = 0
    pos <- x > 0
    2 * sum(x[pos] * log(x[pos] / np0[pos]))
  }
  df <- k - 1
  pval_asym <- 1 - pchisq(stat, df = df)

  # optional Monte Carlo p-value
  pval_mc <- NA_real_
  if (exact) {
    T_obs <- stat
    T_rep <- replicate(B, {
      xr <- as.numeric(rmultinom(1, n, p0))
      if (method == "pearson") sum((xr - np0)^2 / np0)
      else {
        pos <- xr > 0
        2 * sum(xr[pos] * log(xr[pos] / np0[pos]))
      }
    })
    pval_mc <- mean(T_rep >= T_obs)
  }
  list(statistic = stat, df = df,
       p_value = pval_asym, p_value_MC = pval_mc,
       method = method)
}
```

The function returns the statistic, degrees of freedom, and asymptotic
p-value (and optionally an exact Monte Carlo value with globally set seed).

## Task 3 — PYT_1 response distribution in the Product Department (PD)

Hypothesis $H_0: p = p_0 = (\tfrac15,\dots,\tfrac15)$ (uniform distribution
na 5 kategoriach), $\alpha = 0{,}05$.

```{r zad3}
pd <- dane[dane$DZIAL == "PD", ]
tab_pd <- table(factor(pd$PYT_1, levels = c(-2,-1,0,1,2)))
x_pd <- as.numeric(tab_pd)
p0 <- rep(1/5, 5)

tab2 <- data.frame(
  Category  = etyk_pyt1,
  Count = x_pd,
  `p_hat`    = round(x_pd/sum(x_pd), 3),
  `np0`      = sum(x_pd)/5,
  check.names = FALSE
)
kable(tab2, caption = "Observed counts PYT\\_1 w dziale PD vs. expected under uniformity.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))

res_P  <- p_value_multinom(x_pd, p0, method = "pearson")
res_LR <- p_value_multinom(x_pd, p0, method = "LR")

tab3 <- data.frame(
  Test         = c("Pearson chi-square","LR chi-square"),
  Statistic   = c(res_P$statistic, res_LR$statistic),
  df           = c(res_P$df, res_LR$df),
  `p-value`  = c(res_P$p_value, res_LR$p_value),
  check.names  = FALSE
)
tab3$Statistic <- round(tab3$Statistic, 3)
tab3$`p-value` <- signif(tab3$`p-value`, 3)
kable(tab3, caption = "Test results H0: distribution of PYT\\_1 w dziale PD is uniform.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

**Tables description 2--3.** Table 2 shows a clearly non-uniform response distribution: "satisfied" (`PYT_1 = 1`) dominates, and observed counts differ greatly from expected $np_{0i} = 98/5 = 19.6$. In Table 3 both test statistics are large and p-values are well below $\alpha = 0.05$.

**Conclusion.** We reject $H_0$. The PYT_1 response distribution in the Product Department is not uniform — PD employees give positive responses significantly more often than uniformity would imply.

\newpage

# Part II

## Task 4 — Fisher test: GENDER vs CZY_KIER

```{r zad4}
tab_pk <- table(dane$PLEC, dane$CZY_KIER)
kable(addmargins(tab_pk), caption = "Two-way table GENDER x CZY\\_KIER.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))

ft <- fisher.test(tab_pk)
ft
```

**Interpretation.** $p$-value $= `r signif(ft$p.value,3)`$.
Przy poziomie $\alpha = 0.05$ **`r ifelse(ft$p.value < 0.05,"reject","we do not reject")`**
null hypothesis of independence between GENDER and CZY_KIER. Odds ratio estimate is `r round(ft$estimate,3)` (95% PU:
`r round(ft$conf.int[1],3)`--`r round(ft$conf.int[2],3)`).

**Can we infer that $P(K\,|\,\mathrm{mgr}) = P(M\,|\,\mathrm{mgr})$?**

**No, not directly from Fisher's test.** Fisher's test checks **independence**:
$P(K \cap \mathrm{mgr}) = P(K)\cdot P(\mathrm{mgr})$, equivalently
$P(K\,|\,\mathrm{mgr}) = P(K)$. Only if $P(K) = 0.5$ would independence imply
$P(K\,|\,\mathrm{mgr}) = 0.5 = P(M\,|\,\mathrm{mgr})$.
In the sample $P(K) = `r round(mean(dane$PLEC=="K"),3)`$, so the gender distribution
is not 1:1. For "among managers, is $p_K = p_M$?" use a **single-proportion test**
(binomial or $\chi^2$ with $p_0 = 0.5$):

```{r zad4-binom}
n_kier <- sum(dane$CZY_KIER == "Tak")
k_K    <- sum(dane$CZY_KIER == "Tak" & dane$PLEC == "K")
bt <- binom.test(k_K, n_kier, p = 0.5)
bt
```

$p$-value binomial test `r signif(bt$p.value,3)` — therefore
**`r ifelse(bt$p.value<0.05,"reject","we do not reject")`**
hipotezy $P(K\,|\,\mathrm{kier})=1/2$.

## Task 5 — Freeman--Halton test

The Freeman--Halton test generalizes Fisher's test to $r\times c$ tables; in `R` use `fisher.test(..., simulate.p.value = TRUE)`
(because exact computation is costly for larger tables).

```{r zad5-helper}
FH_test <- function(x, y, B = 20000) {
  tb <- table(x, y)
  ft <- fisher.test(tb, simulate.p.value = TRUE, B = B)
  list(table = tb, p_value = ft$p.value)
}

tasks <- list(
  "a) CZY_KIER ~ WIEK_KAT" = list(dane$CZY_KIER, dane$WIEK_KAT),
  "b) CZY_KIER ~ STAZ"     = list(dane$CZY_KIER, dane$STAZ),
  "c) PYT_2 ~ CZY_KIER"    = list(dane$PYT_2,    dane$CZY_KIER),
  "d) PYT_2 ~ STAZ"        = list(dane$PYT_2,    dane$STAZ),
  "e) PYT_2 ~ PLEC"        = list(dane$PYT_2,    dane$PLEC),
  "f) PYT_2 ~ WIEK_KAT"    = list(dane$PYT_2,    dane$WIEK_KAT),
  "c') CZY_ZADOW ~ CZY_KIER" = list(dane$CZY_ZADOW, dane$CZY_KIER),
  "d') CZY_ZADOW ~ STAZ"     = list(dane$CZY_ZADOW, dane$STAZ),
  "e') CZY_ZADOW ~ PLEC"     = list(dane$CZY_ZADOW, dane$PLEC),
  "f') CZY_ZADOW ~ WIEK_KAT" = list(dane$CZY_ZADOW, dane$WIEK_KAT)
)

set.seed(2025)
wyniki <- sapply(tasks, function(z) FH_test(z[[1]], z[[2]])$p_value)
tab5 <- data.frame(
  Hypothesis    = names(wyniki),
  `p-value` = signif(wyniki, 3),
  Decision     = ifelse(wyniki < 0.05, "reject H0", "insufficient evidence"),
  check.names = FALSE
)
kable(tab5, caption = "Freeman-Halton test results (simulation, B = 20000).,
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

**Table description 5.** The *p-value* column contains simulation-based p-values;
the last column gives the decision at $\alpha = 0.05$.

**Comparison PYT_2 vs CZY_ZADOW.** `PYT_2` has 4 categories
(stronger discriminating power), while `CZY_ZADOW` is a binary aggregation of
`PYT_1`. Note:

- for PYT_2 vs (CZY_KIER, TENURE, GENDER, WIEK_KAT) distribution differentiation is usually stronger (more categories — more degrees of freedom, finer dependence picture);
- replacing PYT_2 with CZY_ZADOW aggregates categories and the test usually loses power (p-values rise); for contextual opinions, the full scale is preferable.

## Task 6 — $\chi^2$ independence test: PYT_2 vs CZY_KIER

```{r zad6}
tab_62 <- table(PYT_2 = factor(dane$PYT_2, levels=c(-2,-1,1,2)),
                CZY_KIER = dane$CZY_KIER)
kable(addmargins(tab_62),
      caption = "Two-way table PYT\\_2 x CZY\\_KIER.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))

chi6 <- suppressWarnings(chisq.test(tab_62))
chi6
```

P-value $= `r signif(chi6$p.value,3)`$. Przy $\alpha = 0{,}01$
**`r ifelse(chi6$p.value < 0.01,"reject","we do not reject")`**
independence hypothesis. Result agrees with the Freemana--Haltona
in task 5.

**Association plot (Pearson standardized residuals)** shows the sign and
magnitude of each cell's contribution to $\chi^2$:

```{r zad6-assoc, fig.cap="Association plot: rectangle height proportional to standardized residual; blue = positive, red = negative deviation from independence."}
assoc(tab_62, shade = TRUE, main = NULL)
```

**Figure description 2.** Rectangles above the zero line (blue) are cells with more cases than independence predicts; below (red) — fewer. Largest deviations occur for extreme PYT_2 categories ("str. agree"/"str. disagree") among managers, suggesting opinions on training fit are related to position.

\newpage

## Task 7 — Independence test based on the likelihood ratio

For an $r\times c$ table with counts $n_{ij}$ and total $n$, the statistic:
$$
G^2 = 2 \sum_{i,j} n_{ij} \log\!\Bigl(\frac{n_{ij}}{\hat m_{ij}}\Bigr),
\qquad \hat m_{ij} = \frac{n_{i\cdot}\, n_{\cdot j}}{n},
$$
under $H_0$ of independence is asymptotically $\chi^2_{(r-1)(c-1)}$.

```{r zad7}
LR_indep_test <- function(tab) {
  stopifnot(all(tab >= 0))
  n  <- sum(tab)
  r  <- nrow(tab); c <- ncol(tab)
  mi <- outer(rowSums(tab), colSums(tab)) / n
  pos <- tab > 0
  G2 <- 2 * sum(tab[pos] * log(tab[pos] / mi[pos]))
  df <- (r - 1) * (c - 1)
  list(statistic = G2, df = df, p_value = 1 - pchisq(G2, df))
}

lr <- LR_indep_test(tab_62)
tab7 <- data.frame(
  Test         = c("Pearson chi-square (task 6)", "likelihood ratio (LR)"),
  Statistic   = c(chi6$statistic, lr$statistic),
  df           = c(chi6$parameter, lr$df),
  `p-value`  = c(chi6$p.value, lr$p_value),
  check.names  = FALSE
)
tab7$Statistic <- round(tab7$Statistic, 3)
tab7$`p-value` <- signif(tab7$`p-value`, 3)
kable(tab7, caption = "Comparison of Pearson and LR tests dla PYT\\_2 vs CZY\\_KIER.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

**Table description 7.** Asymptotically $\chi^2$ and $G^2$ share the same limiting distribution, but finite samples can differ — especially with very small cells (PYT_2 = 1 has only 2 observations). Here:
$\chi^2 \approx `r round(chi6$statistic,2)`$ ($p \approx `r signif(chi6$p.value,2)`$),
$G^2 \approx `r round(lr$statistic,2)`$ ($p \approx `r signif(lr$p_value,2)`$).
At $\alpha = 0.05$ both reject $H_0$,
**but at $\alpha = 0.01$ Pearson rejects while LR does not**. The discrepancy comes from low cell counts (`chisq.test` warning); then Fisher/Freeman--Halton (task 5c) with exact or simulated p-values is recommended.

---

# Part III

## Task 8 — Smoking and mortality (lung cancer and heart disease)

Data (annual death rates):

| Disease               | Smokers $\pi_1$ | Non-smokers $\pi_2$ |
|-----------------------|---------------:|------------------:|
| lung cancer              |        0,00140 |           0,00010 |
| coronary heart disease |        0,00669 |           0,00413 |

Three association measures (smoking = exposure, disease/death = outcome):

- proportion difference $\Delta = \pi_1 - \pi_2$,
- relative risk $\mathrm{RR} = \pi_1/\pi_2$,
- odds ratio $\mathrm{OR} = \frac{\pi_1/(1-\pi_1)}{\pi_2/(1-\pi_2)}$.

```{r zad8}
miary <- function(p1, p2) {
  diff <- p1 - p2
  RR   <- p1 / p2
  OR   <- (p1 / (1 - p1)) / (p2 / (1 - p2))
  c(`π1-π2` = diff, RR = RR, OR = OR)
}
lung  <- miary(0.00140, 0.00010)
heart <- miary(0.00669, 0.00413)

tab8 <- rbind(`lung cancer` = lung, `heart disease` = heart)
tab8 <- round(tab8, 5)
kable(tab8, caption = "Association measures: smoking vs death from lung cancer / heart disease.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

**Interpretation.**

- **Lung cancer**: $\mathrm{RR} = 14.0$ means death risk from lung cancer among smokers is **14 times higher** than among non-smokers; OR ≈ RR (because $\pi_i \ll 1$). Proportion difference is small (0.0013), but baseline rate ($\pi_2 \sim 10^{-4}$) makes absolute difference weakly informative.
- **Heart disease**: $\mathrm{RR} \approx 1.62$, $\mathrm{OR} \approx 1.62$. Smokers' risk is about 62% higher. $\Delta = 0.00256$ — larger in absolute terms than lung cancer (heart disease is much more common).

**Strength of association.** Relative measures (RR, OR) show **smoking–lung cancer association is much stronger** (RR = 14 vs RR $\approx 1.62$), though in absolute terms (proportion difference) smoking $\to$ heart disease has greater population impact.

## Task 9 — Association coefficients (gamma, tau)

Pairs considered:

1. PYT_2 (ordinal) vs CZY_KIER (binary — admissible as
   ordinal),
2. PYT_2 vs TENURE (both ordinal),
3. CZY_KIER vs TENURE.

Goodman--Kruskal gamma:
$\gamma = (C - D)/(C + D)$, where $C, D$ are concordant and discordant pair counts.

```{r zad9}
gamma_tab <- function(tab) {
  g <- DescTools::GoodmanKruskalGamma(tab, conf.level = 0.95)
  tau <- DescTools::KendallTauB(tab, conf.level = 0.95)
  c(gamma = unname(g[1]), gamma_L = unname(g[2]), gamma_U = unname(g[3]),
    tau_b = unname(tau[1]), tau_L = unname(tau[2]), tau_U = unname(tau[3]))
}

# Ordinal encodings
tab_9a <- table(dane$PYT_2,    dane$CZY_KIER)
tab_9b <- table(dane$PYT_2,    dane$STAZ)
tab_9c <- table(dane$CZY_KIER, dane$STAZ)

tab9 <- rbind(
  `PYT_2 vs CZY_KIER` = gamma_tab(tab_9a),
  `PYT_2 vs TENURE`     = gamma_tab(tab_9b),
  `CZY_KIER vs TENURE`  = gamma_tab(tab_9c)
)
tab9 <- round(tab9, 3)
kable(tab9,
      caption = "Goodman-Kruskal gamma i Kendall tau-b z 95\\% PU.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

**Table description 9.** Values $\gamma \in [-1,1]$; sign indicates direction of monotonic association, magnitude its strength. Similarly for $\tau_b$ (tie-corrected). Interpretation:

- **PYT_2 vs CZY_KIER** — gamma away from 0 indicates monotonic association (consistent with task 6),
- **PYT_2 vs TENURE** — strength near 0 suggests training-fit opinions do not align monotonically with tenure,
- **CZY_KIER vs TENURE** — directional alignment of management role with tenure.

Confidence intervals containing 0 indicate non-significance of the association measure.

\newpage

## Task 10 — Correspondence analysis "from scratch"

Correspondence analysis (CA) is a graphical method for exploring two-way tables. Let $N$ be a table of dimensions $r\times c$, $n = \sum n_{ij}$,
relative frequency matrix $P = N/n$, mass vectors $r = P\mathbf 1$,
$c = P^\top\mathbf 1$. Standardized residual matrix:
$$
S = D_r^{-1/2} (P - r c^\top) D_c^{-1/2},
\qquad D_r = \mathrm{diag}(r),\ D_c = \mathrm{diag}(c).
$$
SVD: $S = U \Sigma V^\top$. Principal coordinates:
rows $F = D_r^{-1/2} U \Sigma$, columns $G = D_c^{-1/2} V \Sigma$.

```{r zad10-fun}
moja_CA <- function(N, dim = 2) {
  N <- as.matrix(N)
  stopifnot(all(N >= 0))
  n <- sum(N)
  P <- N / n
  r <- rowSums(P); c <- colSums(P)
  Dr_inv_half <- diag(1 / sqrt(r))
  Dc_inv_half <- diag(1 / sqrt(c))
  S <- Dr_inv_half %*% (P - r %o% c) %*% Dc_inv_half
  sv <- svd(S)
  inertia <- sv$d^2
  Fcoord <- Dr_inv_half %*% sv$u %*% diag(sv$d)
  Gcoord <- Dc_inv_half %*% sv$v %*% diag(sv$d)
  rownames(Fcoord) <- rownames(N); rownames(Gcoord) <- colnames(N)
  list(row_coords = Fcoord[, 1:dim, drop = FALSE],
       col_coords = Gcoord[, 1:dim, drop = FALSE],
       singular_values = sv$d,
       inertia = inertia,
       explained = inertia / sum(inertia),
       r_mass = r, c_mass = c, S = S)
}

plot_CA <- function(ca, main = "Correspondence analysis") {
  R <- ca$row_coords; G <- ca$col_coords
  ran <- range(c(R[,1], G[,1], R[,2], G[,2])) * 1.15
  plot(NA, xlim = ran, ylim = ran, xlab = "Dim 1", ylab = "Dim 2",
       main = main, asp = 1)
  abline(h = 0, v = 0, lty = 3, col = "gray60")
  points(R, pch = 19, col = "steelblue")
  text(R, rownames(R), pos = 3, cex = 0.9, col = "steelblue")
  points(G, pch = 17, col = "firebrick")
  text(G, rownames(G), pos = 3, cex = 0.9, col = "firebrick")
  legend("topright", legend = c("rows (PYT_2)", "columns (TENURE)"),
         pch = c(19,17), col = c("steelblue","firebrick"), bty = "n")
}
```

Application to the pair PYT_2 × TENURE:

```{r zad10-apply, fig.cap="Two-dimensional correspondence analysis map for PYT\\_2 (blue) and TENURE (red). Nearby points indicate associated categories."}
tabela_10 <- table(PYT_2 = factor(dane$PYT_2, levels=c(-2,-1,1,2),
                                  labels = etyk_pyt2),
                   STAZ  = factor(dane$STAZ, levels=c(1,2,3),
                                  labels=c("tenure 1","tenure 2","tenure 3")))
ca <- moja_CA(tabela_10)

tab10 <- data.frame(
  Dimension = seq_along(ca$singular_values),
  `sigma (singular value)` = round(ca$singular_values, 4),
  inertia = round(ca$inertia, 4),
  `% explained` = round(ca$explained * 100, 2),
  check.names = FALSE
)
kable(tab10, caption = "Singular values and inertia from CA.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))

plot_CA(ca)
```

**Table description 10 and Figure 3.** Combined inertia of the first two dimensions explains
$`r round(sum(ca$explained[1:2])*100,1)`\%$ of variation. Nearby points indicate strongly associated categories
(e.g. a tenure level co-occurs with a PYT_2 response more than chance). The first axis usually contrasts positive vs negative opinions; the second reveals non-monotonic patterns.

---

# Additional tasks

## Task *1 — Distance correlation test (dCor)

Székely--Rizzo distance correlation: $\mathrm{dCor}(X,Y) = 0 \iff X \perp Y$.
Permutation test p-value:

```{r zad-dcor}
dcor_test <- function(x, y, R = 999) {
  stopifnot(length(x) == length(y))
  obs <- energy::dcor(x, y)
  T_rep <- replicate(R, energy::dcor(x, sample(y)))
  p_val <- (1 + sum(T_rep >= obs)) / (R + 1)
  list(statistic = obs, p_value = p_val, R = R)
}

# Examples
set.seed(2025)
x <- rnorm(200); y1 <- rnorm(200); y2 <- x^2 + rnorm(200, sd = 0.1)

t1 <- dcor_test(x, y1)     # independent
t2 <- dcor_test(x, y2)     # nonlinear dependence

rbind(independent = unlist(t1),
      nonlinear   = unlist(t2)) |>
  round(4) |>
  kable(caption = "Independence test based on dCor (999 permutations).,
        booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

**Table description *1.** For independent $X, Y_1$, dCor is near 0 with large p-value. For nonlinear $Y_2 = X^2$ (undetected by Pearson), dCor detects strong dependence (p-value $\ll 0.05$).

## Task *2 — $|\mathrm{RR} - 1| \le |\mathrm{OR} - 1|$ (RR closer to 1 than OR)

Dla $\pi_1, \pi_2 \in (0,1)$:
$$
\mathrm{OR} - 1 = \frac{\pi_1(1 - \pi_2) - \pi_2(1 - \pi_1)}{\pi_2(1-\pi_1)}
               = \frac{\pi_1 - \pi_2}{\pi_2(1-\pi_1)},
$$
$$
\mathrm{RR} - 1 = \frac{\pi_1 - \pi_2}{\pi_2}.
$$
Stosunek:
$$
\frac{\mathrm{OR} - 1}{\mathrm{RR} - 1} = \frac{1}{1 - \pi_1} \ge 1,
$$
because $1-\pi_1 \in (0,1]$. Both expressions have the same sign (wyznacznik
$\pi_1 - \pi_2$). Hence:
$$
|\mathrm{OR} - 1| \ge |\mathrm{RR} - 1|,
$$
so **RR is not farther from 1 than OR**; equality holds as $\pi_1 \to 0$.

Numerical illustration:

```{r zad-gwiazdka2}
pi1 <- seq(0.01, 0.9, by = 0.05)
pi2 <- 0.1
OR <- (pi1*(1-pi2))/(pi2*(1-pi1))
RR <- pi1/pi2
data.frame(pi1, RR, OR,
           `|RR-1|` = abs(RR-1), `|OR-1|` = abs(OR-1),
           check.names = FALSE) |>
  head(6) |> round(3) |>
  kable(caption = "Illustration: |OR-1| >= |RR-1|.", booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

## Task *3 — Attributable risk

**(a)** $\mathrm{AR} = [P(D) - P(D|E')]/P(D)$ is the
**fraction of cases in the population attributable to exposure $E$**.
The numerator is population risk increase vs. nobody exposed ($P(D|E')$). Dividing by $P(D)$ gives share of total disease burden.

**(b)** From the total probability formula:
$$
P(D) = P(E)P(D|E) + P(E')P(D|E').
$$
Oznaczmy $\pi_E = P(D|E),\ \pi_{E'}=P(D|E'),\ q = P(E)$. Wtedy
$P(D) = q\pi_E + (1-q)\pi_{E'}$, hence:
$$
P(D) - P(D|E') = q(\pi_E - \pi_{E'})
= q\,\pi_{E'}\,(\mathrm{RR} - 1),\qquad \mathrm{RR} = \pi_E/\pi_{E'}.
$$
Dividing by $P(D) = \pi_{E'}(1 + q(\mathrm{RR}-1))$:
$$
\mathrm{AR}
= \frac{q\,\pi_{E'}(\mathrm{RR}-1)}{\pi_{E'}\bigl(1 + q(\mathrm{RR}-1)\bigr)}
= \frac{P(E)(\mathrm{RR}-1)}{1 + P(E)(\mathrm{RR}-1)},
$$
which completes the proof. $\blacksquare$

---

# Summary

- Simultaneous 95% CIs for the PYT_1 response probability vector show clear dominance of the "satisfied" category.
- The `p_value_multinom()` function efficiently tests fit to $p_0$; PYT_1 in PD is not uniform.
- For PYT_2 × CZY_KIER, Fisher/Freeman--Halton, Pearson $\chi^2$, and LR agree: significant association exists. Aggregating PYT_2 to CZY_ZADOW usually reduces power.
- Smoking–lung cancer association is much stronger (RR $\approx 14$) than smoking–heart disease (RR $\approx 1.62$).
- Gamma/tau coefficients and the CA map provide exploratory tools consistent with formal tests.