---
title: "Survey Data Analysis — Report 3"
author: "Michał Marchwiak 276003, Weronika Mitulska 277475"
date: today
lang: en
format:
  pdf:
    pdf-engine: xelatex
    toc: true
    toc-depth: 2
    number-sections: true
    documentclass: article
    geometry:
      - margin=2.2cm
    fontsize: 11pt
    fig-width: 6.5
    fig-height: 4.2
    fig-align: center
    include-in-header:
      text: |
        
        \usepackage{booktabs}
        \usepackage{float}
        \usepackage{amsmath, amssymb}
execute:
  echo: false
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    comment: "#>"
    collapse: true
    dev: "ragg_png"
    dpi: 300
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
  library(kableExtra)
})

set.seed(2025)

# Load data (CSV, separator ';', UTF-8 encoding)
dane <- read.csv2("ankieta.csv", stringsAsFactors = FALSE, fileEncoding = "UTF-8")
colnames(dane) <- c("DZIAL","STAZ","CZY_KIER","PYT_1","PYT_2","PYT_3","PLEC","WIEK")

# Helper variables used in several tasks
dane$CZY_KIER <- factor(dane$CZY_KIER, levels = c("Nie","Tak"))
dane$PYT_2f   <- factor(dane$PYT_2, levels = c(-2,-1,1,2))
dane$STAZf    <- factor(dane$STAZ,  levels = c(1,2,3),
                        labels = c("<1 year","1-3 years",">3 years"))

# "Training satisfaction" variables for two periods (task 3).
# PYT_2 -> first period, PYT_3 -> second period; neither variable
# takes value 0, so binarization relative to zero is unambiguous:
# positive answer (1,2) = "Tak" (satisfied), negative (-2,-1) = "Nie".
dane$CZY_ZADW   <- factor(ifelse(dane$PYT_2 > 0, "Tak", "Nie"), levels = c("Nie","Tak"))
dane$CZY_ZADW_2 <- factor(ifelse(dane$PYT_3 > 0, "Tak", "Nie"), levels = c("Nie","Tak"))
```

# Introduction

This report covers symmetry tests in square tables, Simpson's paradox,
and log-linear models. Some calculations use data from the first task list
(employee survey, $n = 200$, departments IT, HR, MK, PD); others use
data given directly in the task statements.

Survey variable coding conventions:

- **CZY\_KIER** — whether the employee holds a management role (Nie/Tak);
- **PYT\_2** — response to question 2 on ordinal scale $\{-2,-1,1,2\}$ (no zero in data);
- **TENURE** — three categories: `<1 year`, `1–3 years`, `>3 years`;
- **CZY\_ZADW**, **CZY\_ZADW\_2** — training satisfaction in periods 1 and 2, binarized from `PYT_2`/`PYT_3` (positive $\Rightarrow$ "Tak", negative $\Rightarrow$ "Nie").

All computations were performed in R; the random seed was set to `set.seed(2025)`.

---

# Part I

## Task 1 — Conditional symmetry test for a $2\times2$ table

For a $2\times2$ table, symmetry is $H_0:\ p_{12} = p_{21}$. In the **conditional test** we condition on $m = n_{12} + n_{21}$. Under $H_0$, discordant counts follow $\mathrm{Bin}(m,\tfrac12)$, so
$$
n_{12} \mid m \ \sim\ \mathrm{Bin}\!\left(m,\tfrac12\right).
$$
The two-sided p-value doubles the smaller binomial tail:
$$
p = \min\!\Bigl\{1,\ 2\!\!\sum_{j=0}^{\min(n_{12},n_{21})}\binom{m}{j}\Bigl(\tfrac12\Bigr)^{m}\Bigr\}.
$$

```{r zad1, echo=TRUE}
# Conditional (exact) symmetry test for a 2x2 table.
# Returns two-sided p-value based on Bin(m, 1/2) distribution.
p_warunkowy <- function(n12, n21) {
  m <- n12 + n21
  if (m == 0) return(1)
  k <- min(n12, n21)
  min(1, 2 * pbinom(k, size = m, prob = 0.5))
}
```

This test is an exact (conditional) counterpart of McNemar's test; it does
not use a $\chi^2$ approximation, so it remains valid for small counts.

## Task 2 — Efficacy of two analgesic drugs

Data (drug reaction after one hour; same patient for both drugs):

```{r zad2-tab}
M <- matrix(c(1,5,2,4), nrow = 2, byrow = TRUE,
            dimnames = list("Drug A" = c("Negative","Positive"),
                            "Drug B" = c("Negative","Positive")))
M_marg <- addmargins(M)
rownames(M_marg)[3] <- colnames(M_marg)[3] <- "Total"
kable(M_marg, caption = "Response to drug A (rows) vs drug B (columns).,
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

Off-diagonal: $n_{12}=5$, $n_{21}=2$. $H_0$: equal efficacy.

```{r zad2-test}
mcc <- mcnemar.test(M, correct = TRUE)
pw  <- p_warunkowy(5, 2)
tab_z2 <- data.frame(
  Test         = c("McNemar (with continuity correction)", "conditional (task 1)"),
  Statistic   = c(sprintf("%.4f", unname(mcc$statistic)), "---"),
  `p-value`  = c(signif(mcc$p.value, 4), signif(pw, 4)),
  check.names  = FALSE
)
kable(tab_z2, caption = "Testing the hypothesis of equal drug efficacy.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

**Results.** McNemar: $\chi^2 \approx 0.571$ ($p \approx 0.450$). Conditional: $p \approx 0.453$. Both well above $0.05$.

**Conclusion.** Insufficient evidence to reject $H_0$; small sample ($m=7$) implies low power.

## Task 3 — Symmetry of training satisfaction across two periods

Using `CZY_ZADW` and `CZY_ZADW_2` we build a $2\times2$ table and test symmetry.

```{r zad3}
t3 <- table(`Period 1 (CZY_ZADW)` = dane$CZY_ZADW,
            `Period 2 (CZY_ZADW_2)` = dane$CZY_ZADW_2)
t3_marg <- addmargins(t3)
rownames(t3_marg)[3] <- colnames(t3_marg)[3] <- "Total"
kable(t3_marg,
      caption = "Training satisfaction: period 1 (rows) vs period 2 (columns).,
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))

mcc3 <- mcnemar.test(t3, correct = TRUE)
n12 <- t3["Nie","Tak"]; n21 <- t3["Tak","Nie"]
pw3 <- p_warunkowy(n12, n21)
tab_z3 <- data.frame(
  Test        = c("McNemar (with continuity correction)", "conditional (task 1)"),
  Statistic  = c(sprintf("%.4f", unname(mcc3$statistic)), "---"),
  `p-value` = c(signif(mcc3$p.value, 4), signif(pw3, 4)),
  check.names = FALSE
)
kable(tab_z3, caption = "Symmetry test for training satisfaction.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

```{r zad3-plot, fig.cap="Number of employees changing satisfaction status between periods. Off-diagonal cells (Nie$\\to$Tak and Tak$\\to$Nie) determine the symmetry test."}
df3 <- as.data.frame(t3)
colnames(df3) <- c("Period1","Period2","n")
ggplot(df3, aes(Period1, Period2, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 5) +
  scale_fill_gradient(low = "grey90", high = "steelblue") +
  labs(x = "Period 1 (CZY_ZADW)", y = "Period 2 (CZY_ZADW_2)", fill = "Count") +
  theme_minimal(base_size = 11)
```

**Results.** $n_{12}=20$ (Nie→Tak), $n_{21}=8$ (Tak→Nie). McNemar $p\approx0.038$; conditional $p\approx0.036$. Both below $\alpha=0.05$.

**Conclusion.** **We reject** the symmetry hypothesis. This means that the satisfaction
distribution across the two periods **does not** match a symmetry model — therefore
**training satisfaction changed**. The direction of flows
(20 transitions to "Tak" vs 8 to "Nie") indicates an **increase** in the proportion
of satisfied employees in the second period.

\newpage

## Task 4 — Response symmetry ($5\times5$ table, Bowker test)

Task Table 2 data — same question from $n=200$ employees in periods 1 (rows) and 2 (columns):

```{r zad4-tab}
T2 <- matrix(c(10, 2, 1, 1, 0,
                0,15, 1, 1, 0,
                1, 1,32, 6, 0,
                0, 0, 1,96, 3,
                1, 1, 0, 1,26),
             nrow = 5, byrow = TRUE,
             dimnames = list("Period 1" = c("-2","-1","0","1","2"),
                             "Period 2" = c("-2","-1","0","1","2")))
kable(T2, caption = "Responses in period 1 (rows) vs period 2 (columns).,
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

For $k\times k$ tables, **Bowker's test** generalizes McNemar:
$$
\chi^2_B = \sum_{i<j} \frac{(n_{ij}-n_{ji})^2}{n_{ij}+n_{ji}},
$$
with df equal to the number of pairs $(i,j),\ i<j$.

```{r zad4-test}
# Bowker computed manually: pairs with n_ij + n_ji = 0 are skipped (0/0),
# which reduces the effective degrees of freedom.
k <- nrow(T2); bow <- 0; df_eff <- 0
for (i in 1:(k-1)) for (j in (i+1):k) {
  s <- T2[i,j] + T2[j,i]
  if (s > 0) { bow <- bow + (T2[i,j]-T2[j,i])^2 / s; df_eff <- df_eff + 1 }
}
df_nom <- k*(k-1)/2
tab_z4 <- data.frame(
  Variant      = c("nominal df = 10", "effective df = 9 (after removing 0/0 pairs)"),
  Statistic   = round(bow, 4),
  df           = c(df_nom, df_eff),
  `p-value`  = c(signif(1-pchisq(bow, df_nom), 4), signif(1-pchisq(bow, df_eff), 4)),
  check.names  = FALSE
)
kable(tab_z4, caption = "Bowker symmetry test for a 5×5 table.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

**Zero-count problem.** Category pair $(0,2)$, i.e. cells $(3,5)$ i $(5,3)$,
ma $n_{35}+n_{53} = 0+0 = 0$, leading to division $0/0$ (funkcja
`mcnemar.test` then returns `NaN`). This pair contributes zero; omit and reduce df to $9$ (discussed in additional task 1\*).

**Results.** $\chi^2_B \approx 10.57$; $p\approx0.306$ (df=9) or $0.392$ (df=10). Both $p>0.05$.

**Conclusion.** No evidence against symmetry; **no evidence of change** in company approach rating.

\newpage

# Part II

## Task 5 — Simpson's paradox

We compare treatments A (new) and B (old) overall and by comorbidity.

```{r zad5}
skutecz <- function(popr, brak) popr / (popr + brak)
orf <- function(a, b, c, d) (a*d)/(b*c)   # odds ratio (improvement A:B)

tab5 <- data.frame(
  Group = c("Overall group", "With comorbidities", "Without comorbidities"),
  `Improvement A` = c("117/221", "17/118", "100/103"),
  `Improvement B` = c("177/221", "2/38",   "175/183"),
  `% impr. A` = round(100*c(skutecz(117,104), skutecz(17,101), skutecz(100,3)), 1),
  `% impr. B` = round(100*c(skutecz(177,44),  skutecz(2,36),   skutecz(175,8)), 1),
  `OR (A:B)`  = round(c(orf(117,104,177,44), orf(17,101,2,36), orf(100,3,175,8)), 2),
  check.names = FALSE
)
kable(tab5, caption = "Patients with improvement, improvement rate, and odds ratio for treatment A vs B.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

```{r zad5-plot, fig.cap="Improvement rate w metodach A i B: for the overall group method B appears better, ale w obu podgrupach lepsza jest metoda A — direction reversal indicates Simpson's paradox."}
dfp <- tab5 |>
  dplyr::select(Group, `% impr. A`, `% impr. B`) |>
  tidyr::pivot_longer(-Group, names_to = "Method", values_to = "proc")
dfp$Method <- ifelse(dfp$Method == "% impr. A", "Treatment A", "Treatment B")
dfp$Group <- factor(dfp$Group,
                    levels = c("Overall group","With comorbidities","Without comorbidities"))
ggplot(dfp, aes(Group, proc, fill = Method)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = paste0(proc,"%")),
            position = position_dodge(width = 0.7), vjust = -0.3, size = 3.4) +
  scale_fill_manual(values = c("Treatment A" = "steelblue", "Treatment B" = "firebrick")) +
  labs(x = NULL, y = "Improvement rate [%]") +
  ylim(0, 105) +
  theme_minimal(base_size = 11)
```

**Results.** Overall B looks better (80.1% vs 52.9%); **in both subgroups A is better**.

**Conclusion.** **Yes, Simpson's paradox occurs** — the direction of association
reverses after accounting for a confounding variable (comorbidities).
The cause is uneven risk distribution: treatment A was used more often
for sicker patients (118 of 221 treated with A had comorbidities
vs 38 of 221 treated with B), which worsened its aggregated
outcome. The proper effectiveness assessment is subgroup
analysis, which indicates an advantage for treatment A.

## Task 6 — Interpretation of log-linear models

Variables **1=CZY\_KIER**, **2=PYT\_2**, **3=TENURE**. $\pi_{ijk}$ is cell probability; brackets show retained interactions.

```{r zad6-tab}
z6 <- data.frame(
  Model = c("$[1\\ 3]$", "$[13]$", "$[1\\ 2\\ 3]$",
            "$[12\\ 3]$", "$[12\\ 13]$", "$[1\\ 23]$"),
  Postac = c(
    "$\\pi_{i\\cdot k}=\\pi_{i\\cdot\\cdot}\\,\\pi_{\\cdot\\cdot k}$",
    "$\\pi_{i\\cdot k}$ dowolne",
    "$\\pi_{i\\cdot\\cdot}\\,\\pi_{\\cdot j\\cdot}\\,\\pi_{\\cdot\\cdot k}$",
    "$\\pi_{ij\\cdot}\\,\\pi_{\\cdot\\cdot k}$",
    "$\\pi_{ij\\cdot}\\,\\pi_{i\\cdot k}/\\pi_{i\\cdot\\cdot}$",
    "$\\pi_{i\\cdot\\cdot}\\,\\pi_{\\cdot jk}$"),
  Interpretation = c(
    "\\textbf{Independence of CZY\\_KIER and TENURE} (model for pair 1--3, omitting variable 2). Position does not depend on tenure.",
    "\\textbf{Saturated model for pair 1--3} --- allows any association between CZY\\_KIER and TENURE.",
    "\\textbf{Mutual (complete) independence} of all three variables.",
    "\\textbf{TENURE independent of pair (CZY\\_KIER, PYT\\_2)}; variables 1 and 2 may be associated.",
    "\\textbf{PYT\\_2 and TENURE conditionally independent given CZY\\_KIER} (variable 1 is the common core).",
    "\\textbf{CZY\\_KIER independent of pair (PYT\\_2, TENURE)}; variables 2 and 3 may be associated."),
  check.names = FALSE
)
kbl(z6, format = "latex", booktabs = TRUE, escape = FALSE,
    col.names = c("Model", "Form $\\pi_{ijk}$", "Interpretation"),
    caption = "Interpretation modeli log-liniowych (1 = CZY\\_KIER, 2 = PYT\\_2, 3 = TENURE).") |>
  kable_styling(latex_options = c("HOLD_position"), full_width = FALSE) |>
  column_spec(1, width = "1.5cm") |>
  column_spec(2, width = "3cm") |>
  column_spec(3, width = "9.5cm")
```

**Comment.** Models $[1\ 3]$ and $[13]$ concern a two-way table
(zmienne 1 i 3): first assumes independence, second is saturated.
Three-way models range from most restrictive
(complete independence $[1\ 2\ 3]$), through one variable independent of a pair ($[12\ 3]$, $[1\ 23]$), to conditional independence
($[12\ 13]$, where a common variable links both others but they are not associated with each other).

## Task 7 — Probability estimation (models $[123]$ and $[12\ 23]$)

```{r zad7}
tab3w <- table(CZY_KIER = dane$CZY_KIER, PYT_2 = dane$PYT_2f, STAZ = dane$STAZf)

# Saturated model [123] = empirical frequencies
P_sat <- tab3w / sum(tab3w)
p1_sat <- sum(P_sat["Tak","2",]) / sum(P_sat["Tak",,])    # P(PYT_2=2 | kier)
p2_sat <- sum(P_sat[,, "<1 year"]["Tak",]) / sum(P_sat[,, "<1 year"]) # P(manager | tenure<1)
p3_sat <- sum(P_sat[,, ">3 years"]["Nie",]) / sum(P_sat[,, ">3 years"]) # P(non-manager | tenure>3)

# Model [12 23]: 1 and 3 conditionally independent given 2
m1223 <- loglin(tab3w, list(c(1,2), c(2,3)), fit = TRUE, print = FALSE)
Pf <- m1223$fit / sum(m1223$fit)
p1_m <- sum(Pf["Tak","2",]) / sum(Pf["Tak",,])
p2_m <- sum(Pf[,, "<1 year"]["Tak",]) / sum(Pf[,, "<1 year"])
p3_m <- sum(Pf[,, ">3 years"]["Nie",]) / sum(Pf[,, ">3 years"])

tab7 <- data.frame(
  Probability = c(
    "$P(\\text{PYT\\_2}{=}2 \\mid \\text{manager})$",
    "$P(\\text{manager} \\mid \\text{tenure} < 1\\text{ rok})$",
    "$P(\\text{non-manager} \\mid \\text{tenure} > 3\\text{ lata})$"),
  `Model [123]`    = round(c(p1_sat, p2_sat, p3_sat), 4),
  `Model [12 23]`  = round(c(p1_m, p2_m, p3_m), 4),
  check.names = FALSE
)
kbl(tab7, format = "latex", booktabs = TRUE, escape = FALSE,
    caption = "Probability estimates in the saturated and [12 23] models.") |>
  kable_styling(latex_options = c("HOLD_position"))
```

****Results (saturated model $[123]$ = empirical frequencies).****

- $P(\text{PYT\_2}=2 \mid \text{manager}) \approx 0.481$ — nearly half of managers strongly satisfied;
- $P(\text{manager} \mid \text{tenure}<1\text{ year}) \approx 0.024$;
- $P(\text{non-manager} \mid \text{tenure}>3\text{ years}) \approx 0.526$.

**Results (model $[12\ 23]$).** First probability remains the same ($\approx 0{,}481$), because $[12\ 23]$ preserves full pair dependence (CZY\_KIER, PYT\_2). The other two change because CZY\_KIER–TENURE dependence is filtered through PYT\_2 (conditional independence
1 i 3 przy ustalonym 2): $P(\text{kier}\mid\text{tenure}<1) \approx 0{,}128$
oraz $P(\text{nie kier}\mid\text{tenure}>3) \approx 0{,}778$. The model smooths extreme estimates by borrowing information across PYT\_2 layers.

## Task 8 — Testing independence hypotheses (log-linear models)

Each hypothesis: log-linear model vs saturated model; $G^2$ deviance is $\chi^2$ under $H_0$.

```{r zad8}
df8 <- as.data.frame(tab3w)
m_mut  <- glm(Freq ~ CZY_KIER + PYT_2 + STAZ,            family = poisson, data = df8)  # [1 2 3]
m_jnt  <- glm(Freq ~ CZY_KIER*STAZ + PYT_2,              family = poisson, data = df8)  # [2 13]
m_cnd  <- glm(Freq ~ CZY_KIER*STAZ + PYT_2*STAZ,         family = poisson, data = df8)  # [13 23]

g2 <- function(m) c(round(m$deviance,3), m$df.residual, signif(1-pchisq(m$deviance,m$df.residual),4))
tab8 <- data.frame(
  Hypothesis = c("CZY_KIER, PYT_2, TENURE mutually independent",
               "PYT_2 independent of the pair (CZY_KIER, TENURE)",
               "PYT_2 independent of CZY_KIER given TENURE"),
  Model    = c("[1 2 3]", "[2 13]", "[13 23]"),
  `G²`     = c(g2(m_mut)[1], g2(m_jnt)[1], g2(m_cnd)[1]),
  df       = c(g2(m_mut)[2], g2(m_jnt)[2], g2(m_cnd)[2]),
  `p-value` = c(g2(m_mut)[3], g2(m_jnt)[3], g2(m_cnd)[3]),
  check.names = FALSE
)
kable(tab8, caption = "Likelihood-ratio tests for independence hypotheses.",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position"))
```

**Results and conclusions** (significance level $\alpha = 0{,}05$):

1. **Mutual independence** $[1\ 2\ 3]$: reject $H_0$ ($p\approx0.0006$).
2. **PYT\_2 independent of (CZY\_KIER, TENURE)** $[2\ 13]$: insufficient evidence to reject ($p\approx0.081$).
3. **PYT\_2 independent of CZY\_KIER given TENURE** $[13\ 23]$: insufficient evidence to reject ($p\approx0.845$).

```{r zad8-plot, fig.height=5, fig.cap="PYT\\_2 response distribution in strata by tenure (columns) and management role (rows). Each panel has its own y-axis scale (managers are few)."}
df8plot <- as.data.frame(tab3w)
df8plot$CZY_KIER <- factor(df8plot$CZY_KIER, levels = c("Nie","Tak"),
                           labels = c("Non-manager", "Manager"))
ggplot(df8plot, aes(x = PYT_2, y = Freq, fill = PYT_2)) +
  geom_col(show.legend = FALSE, width = 0.75) +
  geom_text(aes(label = Freq), vjust = -0.3, size = 2.7) +
  facet_grid(CZY_KIER ~ STAZ, scales = "free_y") +
  scale_fill_brewer(palette = "RdYlBu") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = "PYT_2 (response)", y = "Number of observations") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"))
```

**Comment on the figure.** Plot shows response distribution PYT\_2 in six strata (3 tenure categories $\times$ 2 position levels). Distribution shape
PYT\_2 is similar across strata (positive responses dominate), consistent with not rejecting hypotheses 2 and 3. Strong dependence is also visible CZY\_KIER–TENURE: managers appear almost only with longer tenure (panel „Manager" for tenure `<1 year` is nearly empty),
explaining rejection of mutual independence.

\newpage

# Additional tasks

## Task 1* — Exact symmetry test for a table with zeros

Bowker hits $0/0$ for pair $(0,2)$; use **exact conditional symmetry test**.

**Idea.** Under symmetry, condition on pair sums; binomial allocation across pairs.

**Determining the critical level.** P-value is the probability under $H_0$ that the statistic is at least as extreme as observed. It is found by **summing over all configurations**
$\{n_{ij}': n_{ij}'+n_{ji}'=m_{ij}\}$ their joint binomial probability (product over pairs), including only layouts where statistic $\ge$ observed value. Pairs with $m_{ij}=0$ are **omitted** (they add no randomness), which naturally removes the $0/0$.
Below we estimate this p-value by Monte Carlo.

```{r zad1gw}
exact_sym_mc <- function(T2, B = 20000) {
  k <- nrow(T2)
  pairs <- which(upper.tri(T2), arr.ind = TRUE)
  m  <- T2[cbind(pairs[,1], pairs[,2])] + T2[cbind(pairs[,2], pairs[,1])]
  nij <- T2[cbind(pairs[,1], pairs[,2])]
  use <- m > 0
  Tobs <- sum((2*nij[use] - m[use])^2 / m[use])   # = suma (n_ij - n_ji)^2/(n_ij+n_ji)
  cnt <- 0
  for (b in 1:B) {
    sim <- rbinom(sum(use), m[use], 0.5)
    Tsim <- sum((2*sim - m[use])^2 / m[use])
    if (Tsim >= Tobs - 1e-9) cnt <- cnt + 1
  }
  list(T = Tobs, p = (cnt + 1)/(B + 1), df_eff = sum(use))
}
set.seed(2025)
es <- exact_sym_mc(T2)
```

Exact test ($B=20000$): $T \approx `r round(es$T,3)`$, p $\approx `r signif(es$p,3)`$ — **insufficient evidence to reject symmetry**.

---

# Summary

- **Zad. 1–2.** Warunkowy symmetry test dla tablicy $2\times2$ is the exact McNemar counterpart ($n_{12}\mid m\sim\mathrm{Bin}(m,\tfrac12)$).
  For analgesics ($p\approx0{,}45$) no evidence of different efficacy.
- **Zad. 3.** Training satisfaction **changed** between periods
  ($p\approx0{,}036$) — significantly more people became satisfied.
- **Zad. 4.** Company approach rating **did not change** — test Bowkera
  ($p\approx0{,}31$) does not reject symmetry (zero problem addressed in sat. 1\*).
- **Zad. 5.** Treatment data show **Simpson's paradox**: metoda B
  wins in aggregate but A is better in both subgroups.
- **Zad. 6–7.** Six log-linear models interpreted; probabilities estimated under saturated and $[12\ 23]$.
- **Zad. 8.** Variables CZY\_KIER, PYT\_2, TENURE **are not** mutually independent;
  PYT\_2 is (conditionally) independent of CZY\_KIER given tenure ($p\approx0{,}85$).